The hydra and the rooster

### 2020-07-24 latex/chapter-ordinals.tex
redaction introduction to ordinals

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as rather simple data types in \gallina. So, we will use \emph{ordinal notations}: Let $\alpha$ be some countable ordinal; In \coq{} terms, an ordinal notation for $\alpha$ will be a data type for representing all ordinals below $\alpha$ (excluding $\alpha$) and a bunch of functions for comparing ordinals, doing some arithmetic, inspecting the topology of the order (limits, successors), and proofs of well-foundedness). These functions should be proved correct, with respect to Schütte's model.

@@

@@

In some cases we want to show that two notation systems describe the same segment (for instance $3+\omega$ and $\omega$). For that purpose, one may prove that the two notation systems are order-isomorphic.

\begin{Coqsrc}
Class  ON_Isomorphic 
       `(OA : @ON A ltA compareA)
       `(OB : @ON B ltB  compareB)
       (f : A -> B)
       (g : B -> A):=
  {
  iso_compare_comm :forall x y : A,  compareB (f x) (f y) =
                                 compareA x y;
  iso_inv1 : forall a, g (f a)= a;
  iso_inv2 : forall b, f (g b) = b}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Class ON_for `(alpha : Schutte_basics.Ord)
     `(OA : @ON A ltA  compareA)
      (iota : A -> Schutte_basics.Ord) :=
  { conform_inj : forall a, Schutte_basics.lt (iota a) alpha;
    conform_surj : forall beta, Schutte_basics.lt beta alpha ->
                                exists b, iota b = beta;
    conform_compare : forall a b:A,
        match compareA a b with
          Datatypes.Lt => Schutte_basics.lt (iota a) (iota b)
        | Datatypes.Eq => iota a = iota b
        | Datatypes.Gt => Schutte_basics.lt (iota b) (iota a)
        end}.
\end{Coqsrc}

@@

\subsection{Examples}

The simplest example of ordinal notation is

>>>>

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as rather simple data types in \gallina. So, we will use \emph{ordinal notations} (also called \emph{ordinal notation systems}). Let $\alpha$ be some countable ordinal; In \coq{} terms, an ordinal notation for $\alpha$ will be a data type for representing all ordinals below $\alpha$ (excluding $\alpha$) and a bunch of functions for comparing ordinals, doing some arithmetic, inspecting the topology of the order (limits, successors), and proofs of well-foundedness). These functions should be proved correct, with respect to Schütte's model.

@@

\subsubsection{Examples}

The simplest example of ordinal notation is built over the type \texttt{nat} of \coq's standard library. We have only to apply already proven lemmas about Peano numbers.

\vspace{4pt}
\noindent\emph{From Library~\href{../src/html/hydras.OrdinalNotations/Omega_ON.html}{OrdinalNotations.Omega\_ON}}

\begin{Coqsrc}
Global Instance Omega : ON  Peano.lt Nat.compare.
Proof.
 split.
 - apply Nat.lt_strorder.
 - apply Wf_nat.lt_wf.
 - apply Nat.compare_spec.
Qed.
\end{Coqsrc}

More complex examples will be developped in further sections or chapters. For instance,
a notation system for ordinals less than $\epsilon_0$ is presented in chapter~\ref{chap:T1}.
Notation systems for $\omega+\omega$, $\omega^2$, etc. will be given in the next paragraphs.

@@

In some cases we want to show that two notation systems describe the same segment (for instance $[0,3+\omega)$ and $[0,\omega($). For this purpose, one may prove that the two notation systems are order-isomorphic.

\begin{Coqsrc}
Class  ON_Isomorphic 
       `(OA : @ON A ltA compareA)
       `(OB : @ON B ltB  compareB)
       (f : A -> B)
       (g : B -> A):=
  {
  iso_compare: forall x y : A,  compareB (f x) (f y) =
                                 compareA x y;
  iso_inv1 : forall a, g (f a)= a;
  iso_inv2 : forall b, f (g b) = b}.
\end{Coqsrc}

@@

\begin{Coqsrc}
Class ON_for `(alpha : Schutte_basics.Ord)
     `(OA : @ON A ltA  compareA)
      (iota : A -> Schutte_basics.Ord) :=
  { ON_for_inj : forall a, Schutte_basics.lt (iota a) alpha;
    ON_for_onto : forall beta, Schutte_basics.lt beta alpha ->
                                exists b, iota b = beta;
    On_compare_spec : forall a b:A,
        match compareA a b with
          Datatypes.Lt => Schutte_basics.lt (iota a) (iota b)
        | Datatypes.Eq => iota a = iota b
        | Datatypes.Gt => Schutte_basics.lt (iota b) (iota a)
        end}.
\end{Coqsrc}

@@
